Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-21 manuscript/main.tex
added figures

\section{Introduction}
Suffix array is a data structure used in many fields, such as stringology,
bioinformatics, and... .
Particularly in bioinformatics, it is used as part of FM index, which in turn
is used for read mapping.
Recently, a new paradigm in bioinformatics was formed, moving from a single
reference genome to a pangenome.
The term pangenome was first used by Tettelin et al. in 2005 during the
resequencing experiments on streptococcus.
Since then, the definition of a pangenome has shifted to any set of sequences
analysed jointly.
Nevertheless, most pangenomes consist of highly related sequences and thus are
highly repetitive.
This leads to representation of pangenomes as graphs, where similar genome
regions are represented by a single node and the nodes are conected to paths
representing the original sequences.
Several approaches of pangenomic graph construction exist, such as variation
graphs, cactus graphs, ...
Most of these approaches require an initial local alignment of similar regions,
which makes them computationally expensive.
Here we present a new class of graphs, prefix-free graphs, which are easy to
construct and provide a way how to efficiently index them to allow pattern 
search queries.
Pangenomes introduced by...
In computational pangenomics...
Repetitive datasets...
Graph pangenomes and textual pangenomes...
Prefix-free parsing is a technique...
Prefix-free graphs play an important role in improving space
complexity of several efficient algorithms, but are never the concept is never
framed as a pangenomic data structure.
We think that separating and clearly defining this data structure can lead to
more advancements...
Our contribution is in defining a data structure called prefix-free graphs and
showing how this data structure can be used to generate suffix array in
sublinear space.
Furthermore, we implemented two binarieš for an easy construction of prefix-free
graphs from a set of sequences, and from a graph in gfa format.
Furthermore, we implemented rust library for working with prefix-free graphs.
This library contains an iterator, which can be directly used to generate the 
suffix array in sublinear space.

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{}.
% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
    \label{fig:gfa}
\end{figure}

>>>>

\section{Introduction}
Suffix array is a data structure used in many fields, such as stringology,
bioinformatics, and... .
Particularly in bioinformatics, it is used as part of FM index, which in turn
is used for read mapping.
Recently, a new paradigm in bioinformatics was formed, moving from a single
reference genome to a pangenome.
The term pangenome was first used by \citet{tettelin2005genome} in 2005 during the
resequencing experiments on streptococcus.
Since then, the definition of a pangenome has shifted to any set of sequences
analysed jointly \cite{computational2018computational}.
Nevertheless, most pangenomes consist of highly related sequences and thus are
highly repetitive.
This leads to representation of pangenomes as graphs, where similar genome
regions are represented by a single node and the nodes are conected to paths
representing the original sequences.
Several approaches of pangenomic graph construction exist, such as variation
graphs \cite{church2015extending}?, cactus graphs \cite{paten2011cactus}, 
Wheeler graphs \cite{gagie2017wheeler} ...
Most of these approaches require an initial local alignment of similar regions,
which makes them computationally expensive.
Here we present a new class of graphs, prefix-free graphs, which are easy to
construct and provide a way how to efficiently index them to allow pattern 
search queries.
Pangenomes introduced by...
In computational pangenomics...
Repetitive datasets...
Graph pangenomes and textual pangenomes...
Prefix-free parsing is a technique...
Prefix-free graphs play an important role in improving space
complexity of several efficient algorithms, but are never the concept is never
framed as a pangenomic data structure.
We think that separating and clearly defining this data structure can lead to
more advancements...
Our contribution is in defining a data structure called prefix-free graphs and
showing how this data structure can be used to generate suffix array in
sublinear space.
Furthermore, we implemented two binaries for an easy construction of prefix-free
graphs from a set of sequences, and from a graph in gfa format.
Furthermore, we implemented rust library for working with prefix-free graphs.
This library contains an iterator, which can be directly used to generate the 
suffix array in sublinear space.

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
%% strings into blocks such that long repeated substrings are parsed the same
%% way (except possibly at the beginning or end of the substrings).

% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.

@@

\begin{figure}
    \centering
    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
    \label{fig:gfa}
\end{figure}
